---
title: "Analysis for Kilimanjaro IPBES Workshop Paper"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---

R code for analyzing inter-group differences for: Masao CA, Prescott GW, Snethlage MA, Urbach D, Torre-Marin Rando A, Molina-Venegas R, Mollel NP, Hemp C, Hemp A, Fischer M (2022). Stakeholder Perspectives on Nature, People, and Sustainability at Mount Kilimanjaro. People and Nature.
  
```{r}
library(tidyverse)
library(vegan)
```

For background on Permutational Analysis of Variance (PERMANOVA):
Anderson 2014 Permutational Analysis of Variance (PERMANOVA). Wiley StatsRef: Statistical Reference Online https://onlinelibrary.wiley.com/doi/epdf/10.1002/9781118445112.stat07841

Structure of the analyses:

Three questions based on the Ecosystem Services Questionnaire:

1. Are there significant differences between groups in listed NCPs? (Response ~ Group)

2. Are there significant differences between groups and NCP categories in reported changes in access and provision between 2008-2018? (Response ~ Group * ESCODE)

3. Are there significant differences between groups and NCP categories in predictions of how NCP access and provision will change between 2018-2028? (Response ~ Group * ESCODE)

And four further questions based on the Habitat Questionnaire:

4. Are there significant differences between groups and habitats in the reported changes in habitat area between 2008-2018? (Response ~ Group * Biome2)

5. Are there significant differences between groups and habitats in the reported direct drivers of these changes? (Response ~ Group * Biome2)

6. Are there significant differences between groups and habitats in the reported indirect drivers of these changes? (Response ~ Group * Biome2)

7. Are there significant differences between groups and habitats in what Actions are suggested to ensure a sustainable future for people and nature in Mount Kilimanjaro? (Response ~ Group * Biome2)

Enter data. Data available from: https://zenodo.org/record/5846443

```{r}
# habitat questionnaire
habitat.questionnaire.original <- read.csv(
  "./kilimanjaro_ipbes_workshop_habitat_questionnaire.csv")

# ecosystem services questionnaire 
es.questionnaire.original <- read.csv(
  "./kilimanjaro_ipbes_workshop_ecosystem_services_questionnaire.csv",sep = ";")
```

1. Are there significant differences between groups in listed NCPs? (Response ~ Group)

```{r}
es.wider <- es.questionnaire.original %>%
  mutate(ESCODE = str_replace_all(ESCODE, c("ES.HAB" = "Habitat",
                                            "ES.POL" = "Pollination&PestControl",
                                            "ES.PST" = "Pollination&PestControl",
                                            "ES.AIR" = "Air&Climate",
                                            "ES.CLI" = "Air&Climate",
                                            "ES.OCE" = "Water",
                                            "ES.WQN" = "Water",
                                            "ES.WQL" = "Water",
                                            "ES.SOL" = "Soil&Hazards",
                                            "ES.HAZ" = "Soil&Hazards",
                                            "ES.NRG"= "Energy&Materials",
                                            "ES.MAT"= "Energy&Materials",
                                            "ES.FOD"= "Food&Medicine",
                                            "ES.MED"= "Food&Medicine",
                                            "ES.LRN"= "Cultural",
                                            "ES.EXP"= "Cultural",
                                            "ES.IDE"= "Cultural",
                                            "ES.OPT"= "Cultural",
                                            "ES.LIV" = "Livelihoods",
                                            "ES.WEB" = "Livelihoods"))) %>%
  select(Participant, Group, ESCODE) %>%
  group_by(Participant, Group, ESCODE) %>%
  summarise(Count = length(ESCODE)) %>%
  pivot_wider(names_from = ESCODE, values_from = Count, values_fill = list(Count=0))

es.form.data.response <- es.wider[,-c(1:2)]
es.form.data.predictor <- es.wider[,c(1:2)]

set.seed(42)
adonis(es.form.data.response ~ Group, 
       data = es.form.data.predictor, permutations = 999)
```

2. Are there significant differences between groups and NCP categories in reported changes in access and provision between 2008-2018? (Response ~ Group * ESCODE)

```{r}
ncp.access.wide.format <- es.questionnaire.original %>%
  select(Participant, Group, ESCODE, Provision, Access, Provision.will, Access.will)%>%
  mutate(Provision = str_replace_all(Provision, c("Decrease" = "decrease",
                                          "Increase" = "increase",
                                          "Not" = "not",
                                          "Deteriorate" = "decrease",
                                          "No answer" = "NA",
                                          "no answer" = "NA",
                                          "Mixed" = "NA",
                                          "decreased" = "decrease",
                                          "increased" = "increase",
                                          "not changed" = "Neutral",
                                          "not change" = "Neutral",
                                          "decrease" = "Negative",
                                          "deteriorate" = "Negative",
                                          "improve" = "Positive",
                                          "increase" = "Positive",
                                          "Improve" = "Positive"))) %>%
  mutate(Provision.will = str_replace_all(Provision.will, c("Decrease" = "decrease",
                                          "Increase" = "increase",
                                          "Not" = "not",
                                          "Deteriorate" = "decrease",
                                          "No answer" = "NA",
                                          "no answer" = "NA",
                                          "Mixed" = "NA",
                                          "decreased" = "decrease",
                                          "increased" = "increase",
                                          "not changed" = "Neutral",
                                          "not change" = "Neutral",
                                          "decrease" = "Negative",
                                          "deteriorate" = "Negative",
                                          "improve" = "Positive",
                                          "increase" = "Positive",
                                          "Improve" = "Positive"))) %>%
  mutate(Access = str_replace_all(Access, c("Decrease" = "decrease",
                                          "Increase" = "increase",
                                          "Not" = "not",
                                          "Deteriorate" = "decrease",
                                          "No answer" = "NA",
                                          "no answer" = "NA",
                                          "Mixed" = "NA",
                                          "decreased" = "decrease",
                                          "increased" = "increase",
                                          "not changed" = "Neutral",
                                          "not change" = "Neutral",
                                          "decrease" = "Negative",
                                          "deteriorate" = "Negative",
                                          "improve" = "Positive",
                                          "increase" = "Positive",
                                          "Improve" = "Positive"))) %>%
  mutate(Access.will = str_replace_all(Access.will, c("Decrease" = "decrease",
                                          "Increase" = "increase",
                                          "Not" = "not",
                                          "Deteriorate" = "decrease",
                                          "No answer" = "NA",
                                          "no answer" = "NA",
                                          "Mixed" = "NA",
                                          "decreased" = "decrease",
                                          "increased" = "increase",
                                          "not changed" = "Neutral",
                                          "not change" = "Neutral",
                                          "decrease" = "Negative",
                                          "deteriorate" = "Negative",
                                          "improve" = "Positive",
                                          "increase" = "Positive",
                                          "Improve" = "Positive"))) %>%
  mutate(ESCODE = str_replace_all(ESCODE, c("ES.HAB" = "Habitat",
                                            "ES.POL" = "Poll. & pest cont.",
                                            "ES.PST" = "Poll. & pest cont.",
                                            "ES.AIR" = "Air & Climate",
                                            "ES.CLI" = "Air & Climate",
                                            "ES.OCE" = "Water",
                                            "ES.WQN" = "Water",
                                            "ES.WQL" = "Water",
                                            "ES.SOL" = "Soil & hazards",
                                            "ES.HAZ" = "Soil & hazards",
                                            "ES.NRG"= "Energy & materials",
                                            "ES.MAT"= "Energy & materials",
                                            "ES.FOD"= "Food & medicine",
                                            "ES.MED"= "Food & medicine",
                                            "ES.LRN"= "Cultural",
                                            "ES.EXP"= "Cultural",
                                            "ES.IDE"= "Cultural",
                                            "ES.OPT"= "Cultural",
                                            "ES.LIV" = "Livelihoods",
                                            "ES.WEB" = "Livelihoods"))) %>%
   mutate(Provision = as.numeric(str_replace_all(Provision, c("Positive" = "3",
                                                                       "Negative" = "1",
                                                                       "Neutral" = "2")))) %>%
   mutate(Provision.will = as.numeric(str_replace_all(Provision.will, c("Positive" = "3",
                                                                       "Negative" = "1",
                                                                       "Neutral" = "2")))) %>%
   mutate(Access = as.numeric(str_replace_all(Access, c("Positive" = "3",
                                                                       "Negative" = "1",
                                                                       "Neutral" = "2")))) %>%
   mutate(Access.will = as.numeric(str_replace_all(Access.will, c("Positive" = "3",
                                                                       "Negative" = "1",
                                                                       "Neutral" = "2")))) %>%
  filter(Access != "NA") %>%
  filter(Access.will != "NA") %>%
  filter(Provision != "NA") %>%
  filter(Provision.will != "NA")

ncp.past.trends.data.response <- ncp.access.wide.format[,c(4:5)]
ncp.past.trends.data.predictor <- ncp.access.wide.format[,c(1:3)]

set.seed(42)
adonis(ncp.past.trends.data.response ~ Group * ESCODE, 
       data = ncp.past.trends.data.predictor, permutations = 999, na.rm = TRUE)
```

3. Are there significant differences between groups and NCP categories in predictions of how NCP access and provision will change between 2018-2028? (Response ~ Group * ESCODE)

```{r}
ncp.future.trends.data.response <- ncp.access.wide.format[,c(6:7)]
ncp.future.trends.data.predictor <- ncp.access.wide.format[,c(1:3)]

set.seed(42)
adonis(ncp.future.trends.data.response ~ Group * ESCODE, 
       data = ncp.future.trends.data.predictor, permutations = 999, na.rm = TRUE)
```

4. Are there significant differences between groups and habitats in the reported changes in habitat area between 2008-2018? (Response ~ Group * Biome2)

```{r}
#Function to convert blank cells to NA, from Holger Brandl  https://stackoverflow.com/questions/24172111/change-the-blank-cells-to-na
empty_as_na <- function(x){
  if("factor" %in% class(x)) x <- as.character(x) 
  ifelse(as.character(x)!="", x, NA)
}

biome.trends.reduced <- habitat.questionnaire.original %>%
  select(Group,
         Participant,
         Biome2,
         Habitat.area) %>%
  mutate(Habitat.area = as.numeric(str_replace_all(Habitat.area, c("Increased" = "3",
                                                                   "Decreased" = "1",
                                                                   "Mixed" = "NA",
                                                                   "No answer" = "NA",
                                                                   "Not changed" = "2"))))# %>%
 
biome.trends.reduced.na.omitted <- na.omit(biome.trends.reduced)
biome.trends.reduced.na.omitted.predictor <- biome.trends.reduced.na.omitted[,c(1:3)]
biome.trends.reduced.na.omitted.response <- biome.trends.reduced.na.omitted[,-(1:3)]

set.seed(42)
adonis(biome.trends.reduced.na.omitted.response ~ Group * Biome2, 
       data = biome.trends.reduced.na.omitted.predictor, permutations = 999, na.rm = TRUE)
```

5. Are there significant differences between groups and habitats in the reported direct drivers of these changes? (Response ~ Group * Biome2)

```{r}
habitat.questionnaire.dir <- habitat.questionnaire.original %>%
  select(Group,
         Participant,
         Biome2,
         DIR.ACT,
         DIR.CC,
         DIR.IAS,
         DIR.LUC,
         DIR.OVR,
         DIR.POL) %>%
  filter(DIR.ACT + DIR.CC + DIR.IAS + DIR.LUC + DIR.OVR + DIR.POL > 0)
  
habitat.questionnaire.dir.predictor <- habitat.questionnaire.dir[,c(1:3)]
habitat.questionnaire.dir.response <- habitat.questionnaire.dir[,-c(1:3)] %>%
               mutate_if(is.numeric, ~1 * (. > 0))

set.seed(42)
adonis(habitat.questionnaire.dir.response ~ Group * Biome2, 
       data = habitat.questionnaire.dir.predictor, permutations = 999)
```

6. Are there significant differences between groups and habitats in the reported indirect drivers of these changes? (Response ~ Group * Biome2)

```{r}
habitat.questionnaire.ind <- habitat.questionnaire.original %>%
  select(Group,
         Biome2,
         Participant,
         IND.CLT,
         IND.DEM,
         IND.ECO,
         IND.GVN,
         IND.LUC,
         IND.S.T) %>%
  filter(IND.CLT + IND.DEM + IND.ECO + IND.GVN + IND.LUC + IND.S.T > 0)
  
habitat.questionnaire.ind.predictor <- habitat.questionnaire.ind[,c(1:3)]
habitat.questionnaire.ind.response <- habitat.questionnaire.ind[,-c(1:3)] %>%
               mutate_if(is.numeric, ~1 * (. > 0))

set.seed(42)
adonis(habitat.questionnaire.ind.response ~ Group * Biome2, 
       data = habitat.questionnaire.ind.predictor, permutations = 999)
```

7. Are there significant differences between groups and habitats in what Actions are suggested to ensure a sustainable future for people and nature in Mount Kilimanjaro? (Response ~ Group * Biome2)

```{r}
habitat.questionnaire.act <- habitat.questionnaire.original %>%
  select(Group,
         Participant,
         Biome2,
         ACT.LAND,
         ACT.SPEC,
         ACT.AWAR,
         ACT.ENFORC,
         ACT.ECON,
         ACT.PROTECT,
         ACT.LAWPOL,
         ACT.RSRCH,
         ACT.EDUC,
         ACT.INST) %>%
  filter(ACT.LAND + ACT.SPEC + ACT.AWAR + ACT.ENFORC + ACT.ECON + ACT.PROTECT + ACT.LAWPOL + ACT.RSRCH + ACT.EDUC + ACT.INST > 0)
  
habitat.questionnaire.act.predictor <- habitat.questionnaire.act[,c(1:3)]
habitat.questionnaire.act.response <- habitat.questionnaire.act[,-c(1:3)] %>%
               mutate_if(is.numeric, ~1 * (. > 0))

set.seed(42)
adonis(habitat.questionnaire.act.response ~ Group * Biome2, 
       data = habitat.questionnaire.act.predictor, permutations = 999)
```

Load table with unadjusted p-values

```{r}
permanova.results.consolidated <- read.csv(
  "./p_values_raw_Kilimanjaro_IPBES_workshop_PERMANOVA_setseed42.csv")
```

Apply the Benjamini-Hochberg procedure to the p-values
For further details, see: Benjamini and Hochberg 1995 Controlling the False Discovery Rate: A Practical and Powerful Approach to Multiple Testing. Journal of the Royal Statistical Society: Series B (Methodological).  https://doi.org/10.1111/j.2517-6161.1995.tb02031.x

```{r}
p<- permanova.results.consolidated$p
round(p.adjust(p, "BH"), 3)
```

